---
title: "COPS tutorial, atopic dermatitis (AD)"
author: "Teemu Rintala and Vittorio Fortino"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{COPS tutorial, AD}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style type="text/css">

body{ /* Normal  */
      font-size: 12px;
  }
td {  /* Table  */
  font-size: 10px;
}
h1.title {
  font-size: 30px;
  color: DarkRed;
}
h1 { /* Header 1 */
  font-size: 24px;
  color: DarkBlue;
}
h2 { /* Header 2 */
    font-size: 20px;
  color: DarkBlue;
}
h3 { /* Header 3 */
  font-size: 15px;
  font-family: "Times New Roman", Times, serif;
  color: DarkBlue;
}
code.r{ /* Code block */
    font-size: 10px;
}
pre { /* Code block - determines code spacing between lines */
    font-size: 10px;
}
</style>

## Goals

1. Pre-processing 
    + batch correction
    + normalization
2. Clustering with COPS
3. Comparing approaches

## Load data (remove after including the data in COPS)

```{r, eval = FALSE}
pso
```


# TREAGermany (no batch info)

## Load gene-expression data

In this tutorial we use RNA-Seq data from the TREATGermany cohort on GEO. 

```{r, eval = FALSE}
library(curl)
gex_url <- "https://ftp.ncbi.nlm.nih.gov/geo/series/GSE157nnn/GSE157194/suppl/GSE157194%5FRaw%5Fgene%5Fcounts%5Fmatrix%2Etxt%2Egz"
gex_zip <- gzcon(curl(gex_url))
gex_raw <- textConnection(readLines(gex_zip))
close(gex_zip)
gex <- read.table(gex_raw, header = TRUE, row.names = 1, sep = "\t")
close(gex_raw)
```

## Load clinical data

Clinical data includes info on which treatment was used on which patient (cyclosporine or dupilumab). But otherwise, all the necessary information is encoded in the sample names. 
```{r, eval = FALSE}
cli_url <- "https://ftp.ncbi.nlm.nih.gov/geo/series/GSE157nnn/GSE157194/matrix/GSE157194_series_matrix.txt.gz"
cli_zip <- gzcon(curl(cli_url))
cli_raw <- textConnection(readLines(cli_zip))
close(cli_zip)
cli <- read.table(cli_raw, header = FALSE, row.names = NULL, sep = "\t", skip = 30, fill = TRUE)
close(cli_raw)
```

```{r, eval = FALSE}
cli[[1]] <- gsub("!", "", cli[[1]])
for (i in unique(cli[[1]][duplicated(cli[[1]])])) {
  ind <- which(cli[[1]] == i)
  cli[[1]][ind] <- paste(cli[[1]][ind], 1:length(ind), sep = "_")
}

cli <- as.data.frame(t(cli))
colnames(cli) <- cli[1,]
rownames(cli) <- cli[["Sample_title"]]
cli <- cli[-1,]
```

```{r, eval = FALSE}
lapply(cli, table, useNA = "always")
```

## Extract clinical info from sample names
```{r, eval = FALSE}

```


